---
title: 
output: 
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
---
<br>

# Comparing Subway and Bus Ridership with Regression Modeling 

"First we need to propose a regression model for subway ridership in 2020.
As the outcome (subway ridership) is continuous, we can fit a linear regression model."

```{r, include=FALSE}
library(tidyverse)
library(patchwork)
library(readr)
library(broom)
library(dbplyr)
library(viridis)
library(reshape2)
library(plotly)
library(lubridate)

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

theme_set(theme_minimal() + theme(legend.position = "bottom"))

#Importing and cleaning data
mta_data = read_csv (file = "./data/MTA_recent_ridership_data_20201123_0.csv",
                     col_types = cols(
                       date = col_date(format = "%mm/%dd/%yy"),
                       `Subways: % Change From 2019 Equivalent Day` = col_number(),
                       `Buses: % Change From 2019 Equivalent Day` = col_number(),
                       `Bridges and Tunnels: % Change From 2019 Equivalent Day` = col_number()
                       ) #only changed the formats of important variables 
) %>%
  janitor::clean_names()
skimr::skim(mta_data)
mta_data =
  mta_data %>%
  subset(select = -c(lirr_total_estimated_ridership, lirr_percent_change_from_2019_monthly_weekday_saturday_sunday_average, metro_north_total_estimated_ridership, metro_north_percent_change_from_2019_monthly_weekday_saturday_sunday_average, access_a_ride_total_scheduled_trips, access_a_ride_percent_change_from_2019_monthly_weekday_saturday_sunday_average, bridges_and_tunnels_total_traffic, bridges_and_tunnels_percent_change_from_2019_equivalent_day))
#exclude data for lirr, metronorth, access-a-ride, bridges & tunnel

mta_data = mta_data %>%
  mutate( 
    'subway_2019' = subways_total_estimated_ridership/(1+(subways_percent_change_from_2019_equivalent_day/100)),
    'bus_2019'=
      buses_total_estimated_ridership/(1+(buses_percent_change_from_2019_equivalent_day/100))
    ) %>%
  rename(
    "subway_2020" = subways_total_estimated_ridership,
    "subway_pct_change" = subways_percent_change_from_2019_equivalent_day,
    "bus_2020" = buses_total_estimated_ridership,
    "bus_pct_change" = buses_percent_change_from_2019_equivalent_day
    )

#change date to date format and order by date
plot_mta =
  mta_data %>%
  mutate(
    date= as.Date(date,format = "%m/%d")) %>%
  arrange(date)
#create a text_label label
text_label =
  plot_mta %>%
  mutate(text_label = str_c("percent change from 2019 to 2020: ", subway_pct_change)) %>%
  select(date, text_label)

#separate by month and day
mta_data = 
  mta_data %>%
  arrange(date) %>%
  separate(date, into = c("month", "day", "year"))%>%
  mutate(month = as.numeric(month),
         day = as.numeric(day)) %>%
  select(-c(year)) #drop year column

mta_subway_ridership =
  mta_data %>%
  group_by(month)%>%
  summarize(
    avg_subway_2019 = mean(subway_2019),
    avg_subway_2020 = mean(subway_2020)
  )
mta_subway_ridership%>%knitr::kable()
```

```{r, message=FALSE}
subway_ridership = lm(subway_2020 ~ month, data = mta_data)

subway_ridership %>% 
  broom::tidy() 

#to tidy the output and get only the intercept, slope and p-values 
subway_ridership %>% 
  broom::tidy() %>% 
  select(term, estimate, p.value) %>% 
  knitr::kable(digits = 3)

#to plot for regression line 
  ggplot(mta_data, aes(month, subway_2020)) +
  geom_point() +
  stat_smooth(method = lm)+
   xlab("Month in 2020") +
  ylab("Subway Ridership") +
  ggtitle("Subway Ridership 2020 Regression")
```

Now we will build a plot of model residuals against fitted values for subway ridership in 2020.

```{r, message=FALSE}
mta_data %>% 
  modelr::add_predictions(subway_ridership) %>% 
  modelr::add_residuals(subway_ridership) %>% 
  ggplot(aes(x = pred, y = resid)) + geom_point() + 
  labs(x = "Predicted value", 
       y = "Residual") +
    ggtitle("Residual vs. Predicted value of Subway 2020 Ridership Regression Model")
```

Now, we need to propose a regression model for bus ridership in 2020.

As the outcome (bus ridership) is continuous, we can fit a linear regression model.

```{r, message=FALSE}
bus_ridership = lm(bus_2020 ~ month, data = mta_data)
bus_ridership %>% 
  broom::tidy() 

#Now we need to tidy the output and get only the intercept, slope and p-values 
bus_ridership %>% 
  broom::tidy() %>% 
  select(term, estimate, p.value) %>% 
  knitr::kable(digits = 3)

#to plot for regression line
ggplot(mta_data, aes(month, bus_2020)) +
  geom_point() +
  stat_smooth(method = lm)+
   xlab("Month in 2020") +
  ylab("Bus Ridership") +
  ggtitle("Bus Ridership 2020 Regression")
```


* Next, we will build a plot of model residuals against fitted values for bus ridership in 2020.

```{r, message=FALSE}
mta_data %>% 
  modelr::add_predictions(bus_ridership) %>% 
  modelr::add_residuals(bus_ridership) %>% 
  ggplot(aes(x = pred, y = resid)) + geom_point() + 
  labs(x = "Predicted value", 
       y = "Residual")
ggtitle("Residual vs. Predicted value of Subway 2020 Ridership Regression Model")
```

