---
title: 
output: 
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
---
<br>

# __Data__

### Data Source 

We used a publicly available [dataset](https://new.mta.info/coronavirus/ridership) on day-by-day ridership numbers  from the MTA. The dataset contains information on total estimated ridership and percentage change from 2019 by day, beginning March 1, 2020. Estimates of ridership by subway, bus (local, limited, SBS, and Express), Long Island Rail Road, Metro-North Railroad, Access-A-Ride, and Bridges and Tunnels are all available, however, we decided to restrict our analysis to subway and bus ridership.   

The dataset was downloaded as a .csv files and imported into RStudio. We then cleaned the variables names, verified that there were no missing values, and excluded the variables on the Long Island Rail Road, Metro-North Railroad, Access-A-Ride, and Bridges and Tunnels, which we did not plan to include in further data visualization and analysis. 

### Data Cleaning 

The dataset was downloaded as a .csv file and imported into RStudio. We then cleaned the variables names, verified that there were no missing values, and excluded the variables on the Long Island Rail Road, Metro-North Railroad, Access-A-Ride, and Bridges and Tunnels, which we did not plan to include in further data visualization and analysis. 

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(patchwork)
library(readr)
library(broom)
library(dbplyr)
library(viridis)
library(reshape2)
library(plotly)
library(lubridate)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
)

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

theme_set(theme_minimal() + theme(legend.position = "bottom"))
```


```{r, results=FALSE}
mta_data = read_csv (file = "./data/MTA_recent_ridership_data_20201123_0.csv",
                     col_types = cols(
                       date = col_date(format = "%mm/%dd/%yy"),
                       `Subways: % Change From 2019 Equivalent Day` = col_number(),
                       `Buses: % Change From 2019 Equivalent Day` = col_number(),
                       `Bridges and Tunnels: % Change From 2019 Equivalent Day` = col_number()
                       ) #only changed the formats of important variables 
) %>%
  janitor::clean_names()
skimr::skim(mta_data)
mta_data =
  mta_data %>%
  subset(select = -c(lirr_total_estimated_ridership, lirr_percent_change_from_2019_monthly_weekday_saturday_sunday_average, metro_north_total_estimated_ridership, metro_north_percent_change_from_2019_monthly_weekday_saturday_sunday_average, access_a_ride_total_scheduled_trips, access_a_ride_percent_change_from_2019_monthly_weekday_saturday_sunday_average, bridges_and_tunnels_total_traffic, bridges_and_tunnels_percent_change_from_2019_equivalent_day))
#exclude data for lirr, metronorth, access-a-ride, bridges & tunnel
```

Next, we calculated the 2019 ridership estimates for subway and buses based on the 2020 estimates and percentage change data. The 2019 estimates were calculated by (1) changing the percent change to decimal, (2) adding 1, (3) dividing 2020 ridership by the resulting value from step 2 to produce the 2019 ridership estimate.

```{r}
mta_data = mta_data %>%
  mutate( 
    'subway_2019' = subways_total_estimated_ridership/(1+(subways_percent_change_from_2019_equivalent_day/100)),
    'bus_2019'=
      buses_total_estimated_ridership/(1+(buses_percent_change_from_2019_equivalent_day/100))
    ) %>%
  rename(
    "subway_2020" = subways_total_estimated_ridership,
    "subway_pct_change" = subways_percent_change_from_2019_equivalent_day,
    "bus_2020" = buses_total_estimated_ridership,
    "bus_pct_change" = buses_percent_change_from_2019_equivalent_day
    )
```









