---
title: 
output: 
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
---
<br>

# MTA Ridership Mobility Trends 2019 and 2020

We first implemented an exploratory analysis of subway and bus MTA ridership trends in 2019 and 2020.  

After initial data cleaning, we changed the date from character to date format and then created a text_label dataset to show percent change from 2019-2020 when hovered on the graph. Then, we created a dataset with variables related to subway and pivoted it so it could be correctly graphed by year. We then merged the subway database with the text_label dataset. Then using plotly, we plotted the general trends of subway ridership per day from March - November. To visualize the ridership trends within the context of the COVID-19 pandemic, we added vertical lines that indicate the first diagnosed COVID-19 case in New York City (3/1/2020), and the days NYC cumulative cases hit 100,000 (04/07) and 200,000 (05/26).

```{r include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(patchwork)
library(readr)
library(broom)
library(dbplyr)
library(viridis)
library(reshape2)
library(plotly)
library(lubridate)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
)

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

theme_set(theme_minimal() + theme(legend.position = "bottom"))
```

```{r include=FALSE}
mta_data = read_csv (file = "./data/MTA_recent_ridership_data_20201123_0.csv",
                     col_types = cols(
                       date = col_date(format = "%mm/%dd/%yy"),
                       `Subways: % Change From 2019 Equivalent Day` = col_number(),
                       `Buses: % Change From 2019 Equivalent Day` = col_number(),
                       `Bridges and Tunnels: % Change From 2019 Equivalent Day` = col_number()
                       ) #only changed the formats of important variables 
) %>%
  janitor::clean_names()
skimr::skim(mta_data)
mta_data =
  mta_data %>%
  subset(select = -c(lirr_total_estimated_ridership, lirr_percent_change_from_2019_monthly_weekday_saturday_sunday_average, metro_north_total_estimated_ridership, metro_north_percent_change_from_2019_monthly_weekday_saturday_sunday_average, access_a_ride_total_scheduled_trips, access_a_ride_percent_change_from_2019_monthly_weekday_saturday_sunday_average, bridges_and_tunnels_total_traffic, bridges_and_tunnels_percent_change_from_2019_equivalent_day))
#exclude data for lirr, metronorth, access-a-ride, bridges & tunnel

mta_data = mta_data %>%
  mutate( 
    'subway_2019' = subways_total_estimated_ridership/(1+(subways_percent_change_from_2019_equivalent_day/100)),
    'bus_2019'=
      buses_total_estimated_ridership/(1+(buses_percent_change_from_2019_equivalent_day/100))
    ) %>%
  rename(
    "subway_2020" = subways_total_estimated_ridership,
    "subway_pct_change" = subways_percent_change_from_2019_equivalent_day,
    "bus_2020" = buses_total_estimated_ridership,
    "bus_pct_change" = buses_percent_change_from_2019_equivalent_day
    )
```

```{r results='hide', message=FALSE}
#change date to date format and order by date
mta_data =
  mta_data %>%
  mutate(
    date= as.Date(date,format = "%m/%d")) %>%
  arrange(date)
#create a text_label label
text_label =
  mta_data %>%
  mutate(text_label = str_c("percent change from 2019 to 2020: ", subway_pct_change)) %>%
  select(date, text_label)

#graph for subway 
#pivoting
plot_subway = 
  mta_data %>%
  select(subway_2019, subway_2020, date) %>%
  melt(., id.vars = "date") %>%
#merge based on month 
  merge(text_label, by = "date") %>%
  #plotting
  plot_ly(
    x = ~date, y = ~value, type = "scatter", mode = "markers",
    color = ~variable, text = ~text_label) %>%
  layout (
    title = " Figure 2. Subway Ridership Trends 2019 - 2020",
    xaxis = list(title ="Month/Day", tickformat = "%m/%d"), #drop year
    yaxis = list(title="Ridership"))  %>%
  add_lines(x =as.Date("2020-03-01"), line = list(dash="dot", color = 'red', width=0.5, opacity = 0.5),name = 'First case on 3/1') %>%
  add_lines(x =as.Date("2020-04-07"), line = list(dash="dot", color = 'red', width=0.5, alpha = 0.5),name = '100K cases in NYC on 04/07') %>%
  add_lines(x =as.Date("2020-05-26"), line = list(dash="dot", color = 'red', width=0.5, alpha = 0.5),name = '200K cases in NYC on 05/26')
#graph for bus

#pivoting
plot_bus = 
  mta_data %>%
  select(bus_2019,bus_2020,date) %>%
  melt(., id.vars = "date") %>% 
#merge based on month 
  merge(text_label, by = "date") %>%
#plotting
  plot_ly(
    x = ~date, y = ~value, type = "scatter", mode = "markers",
    color = ~variable, text = ~text_label) %>%
  layout (
    title = " Figure 1. Bus Ridership Trends 2019 - 2020",
    xaxis = list(title ="Month/Day", tickformat = "%m/%d"),
    yaxis = list(title="Ridership")) %>%
  add_lines(x =as.Date("2020-03-01"), line = list(dash="dot", color = 'red', width=0.5, opacity = 0.5),name = 'First case on 3/1') %>%
  add_lines(x =as.Date("2020-04-07"), line = list(dash="dot", color = 'red', width=0.5, alpha = 0.5),name = '100K cases in NYC on 04/07') %>%
  add_lines(x =as.Date("2020-05-26"), line = list(dash="dot", color = 'red', width=0.5, alpha = 0.5),name = '200K cases in NYC on 05/26')

#ggplot for subway
plot_subway_2 = 
  mta_data %>%
  select(subway_2019, subway_2020, date) %>%
  melt(., id.vars = "date") %>%
  ggplot(aes(x=date, y=value, color=variable)) +
  geom_point(alpha=0.5) +
  geom_smooth(se = FALSE)  +
  geom_vline(xintercept=as.Date("2020-03-01"), linetype="dotted", color = 'red')+
  geom_vline(xintercept=as.Date("2020-04-07"), linetype="dotted", color = 'red')+
  geom_vline(xintercept=as.Date("2020-05-26"), linetype="dotted", color = 'red')+
  geom_text(x=as.Date("2020-03-01"), y=6100000, label ="First case", angle=90, vjust = 1.2, size=3,color='black') +
  geom_text(x=as.Date("2020-04-07"), y=6100000, label ='100K cases', angle=90, vjust = 1.2, size=3,color='black')+
  geom_text(x=as.Date("2020-05-26"), y=6100000, label ='200K cases', angle=90, vjust = 1.2, size=3,color='black')
ggplotly(plot_subway_2)

#ggplot for bus
plot_bus_2 = 
  mta_data %>%
  select(bus_2019, bus_2020,date) %>%
  melt(., id.vars = "date") %>% 
  ggplot(aes(x=date, y=value, color=variable)) +
  geom_point(alpha=0.5) +
  geom_smooth(se = FALSE) +  
  geom_vline(xintercept=as.Date("2020-03-01"), linetype="dotted", color = 'red')+
  geom_vline(xintercept=as.Date("2020-04-07"), linetype="dotted", color = 'red')+
  geom_vline(xintercept=as.Date("2020-05-26"), linetype="dotted", color = 'red')+
  geom_text(x=as.Date("2020-03-01"), y=2400000, label ="First case", angle=90, vjust = 1.2, size=3,color='black') +
  geom_text(x=as.Date("2020-04-07"), y=2400000, label ='100K cases', angle=90, vjust = 1.2, size=3,color='black')+
  geom_text(x=as.Date("2020-05-26"), y=2400000, label ='200K cases', angle=90, vjust = 1.2, size=3,color='black')
ggplotly(plot_bus_2)

```


```{r message=FALSE}
#see the two graphs together:
plot_bus
plot_subway

(plot_bus_2+plot_subway_2 + plot_annotation(
  title = 'Figure 3. Bus and Subway Ridership Trends 2019-2020', theme = theme(plot.title = element_text(size = 24)
)))
```

__Highlights:__ 

* __Overall reduction in bus and subway ridership in 2020 compared to 2019__

* __Substantial drop in bus and subway ridership from March to April 2020__

* __Fewer people chose to use the subway as compared to buses after April, 2020__


In Figure 1., which shows trends in bus ridership, there appears to be a continuous drop in ridership between the first COVID-19 case on March 1, 2020 and April 1, 2020. This trend is followed by an increase in ridership in April through July 2020 and then subsequently a broadening of the range of values around August 2020, with some reductions in values and others remaining generally constant. From the 2019 estimates, we also notice many outlying lower values, produced by decreased bus ridership on Thursdays and Fridays. This is likely a result of fewer commuters on those days. 

In Figure 2., which shows trends in subway ridership, there appears to be similarities to the bus ridership trends from Figure 1.. However, there are fewer overlapping values when comparing the years 2019 and 2020. The increase in subway ridership, beginning around April 2020 is also less extreme than the increase in bus ridership and there is a more narrow range of values in the 2020 subway ridership as compared to bus ridership. Finally, the 2020 subway ridership does not appear to be decreasing in recent months as with bus ridership. These trends suggest that as people began to use public transportation again during the pandemic, fewer people chose to use the subway as compared to buses. 
 
From Figure 3., we can see that the ridership values in March 2020 are similar to those of ridership in March 2019 for both buses and subways. However, from mid-March onward there are overall reduced ridership trends in 2020 as compared to 2019 for both forms of transit. 

